A sample workflow

### 2025-04-25 analyses/african_mammals_workflow.qmd
some small progress

*Hi Victor, it's Lizzie. I will just put my notes to you in italics. You really should remove them. In general, I suggest you focus on getting some code/plots and basic text done that you then share next week with everyone.*

@@

y_i ~ normal(mu, sigma)

mu = alpha[sp[pop]] + beta[sp[pop]]*year

*Need to finish the notation as I always have to look up the nesting notation.*

@@

*Obviously we can charge forward with the full model if you want, I just thought this would be (a) easier for us, (b) way better to show as it's how you should do it and (c) related to a, I suspect JD can be more help, though any country with a good amount of data could work.* 
So here we refine our question to:

What is trend in mammal species population numbers in Africa, both overall and how does it vary acrosss species and populations? 
**Note**: we will actually focus on African mammals with 10+ populations with 5+ observations per population.

### Simulating data
Now we start to simulate!
*Sorry, I did not have time to do this, but I would simulate from the above model or some other simple one you come up with. I would simulate really NICE data, not realistic data.*

@@

nspecies <- 9
npops_persp <- round(runif(nspecies, 2, 15)) # no. of different populations per species
npops <- sum(npops_persp) # total no. of different populations
nobs_perpop <- round(runif(npops, 10, 40)) # no. of different observations per population

@@

# nested intercepts
mu_species_alpha <- rnorm(n = nspecies, mean = log(200), sd = log(5))
sig_species_alpha <- rnorm(n = nspecies, mean = 5, sd = 5)
mu_pop_alpha <- rnorm(n = npops, mean = mu_species_alpha[spid_perpop])
sig_pop_alpha <- abs(rnorm(n = npops))
alpha <- rnorm(n = npops, mean = mu_pop_alpha, sd = sig_pop_alpha)

# nested slopes
mu_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.01)
sig_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.05)
mu_pop_beta <- rnorm(n = npops, mean = mu_species_beta[spid_perpop], sd = 0.05)
sig_pop_beta <- abs(rnorm(n = npops, mean = 0, sd = 0.01))
beta <- rnorm(n = npops, mean = mu_pop_beta, sd = sig_pop_beta)

@@

sigma_obs <- abs(rnorm(1, mean = 0.25, sd = 0.25))

@@

```{r inference sim data}
#| eval: false
#| echo: false
#| 
species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("species", sp), "Std. Error"]}))

@@

@@

ggplot(data = lpi_subset_rsh) +
  geom_line(aes(x = year, y = value, 
                group = paste0(genus_species, ID),
                color = genus_species)) +
  theme_bw()

ggplot(data = lpi_subset_rsh[lpi_subset_rsh$value != 0,]) +
  geom_line(aes(x = year, y = log(value), 
                group = paste0(genus_species, ID),
                color = genus_species)) +
  theme_bw()

@@

runmodel <- TRUE

lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,]
lpi_subset_rsh$year <- lpi_subset_rsh$year-1980
lpi_subset_rsh$logvalue <- log(lpi_subset_rsh$value)
lpi_subset_rsh$ID <- as>character(lpi_subset_rsh$ID)

if(runmodel){
  nestlm <- lm(logvalue ~ genus_species/ID + year * genus_species/ID, data = lpi_subset_rsh)

  saveRDS(nestlm, file = file.path(wd, "output/fit/nestmodel1.rds"))
}else{
  nestlm <- readRDS(file.path(wd, "output/fit/nestmodel1.rds"))
}

coefs <- summary(nestlm)$coefficients

species <- unique(lpi_subset_rsh$genus_species)[order(unique(lpi_subset_rsh$genus_species))]

species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(species[2:length(species)], function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("genus_species", sp), "Estimate"]}))
species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(species[2:length(species)], function(sp){coefs[paste0("genus_species", sp), "Std. Error"]}))

intercepts <- data.frame(est=species_intercept, est.sd=species_intercept_sd, species)

species_slope <- c(coefs["year", "Estimate"], sapply(species[2:length(species)], function(sp){coefs["year", "Estimate"] + coefs[paste0("year:genus_species", sp), "Estimate"]}))
species_slope_sd <- c(coefs["year", "Std. Error"], sapply(species[2:length(species)], function(sp){coefs[paste0("year:genus_species", sp), "Std. Error"]}))

slopes <- data.frame(est=species_slope, est.sd=species_slope_sd, species)

@@

```{r}
#| fig-height: 10
#| fig-width: 12
#| eval: false

retrodata <- data.frame(obs = lpi_subset_rsh$logvalue, pred = predict(nestlm),
                        year = lpi_subset_rsh$year, units = lpi_subset_rsh$Units,
                        species = lpi_subset_rsh$genus_species, id = lpi_subset_rsh$ID)

ggplot(data = retrodata) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  facet_wrap(~ species, scales = "free") +
  geom_point(aes(x = obs, y = pred, group = id, color = as.character(id)),
             size = 0.5) + 
  theme_bw() + theme(legend.position = 'none')

@@

```{r}
#| fig-height: 8
#| fig-width: 8
#| eval: false

ggplot(data = retrodata[retrodata$species == "Loxodonta africana", ]) +
  facet_wrap(~ id, scales = "free") +
  geom_point(aes(x = year, y = obs, group = id, color = id),
             size = 1) + 
  geom_line(aes(x = year, y = pred, group = id, color = id)) +
  theme_bw() + theme(legend.position = 'none')

>>>>

@@

$$\log(y_i) \sim \text{normal}(\mu_i, \sigma)$$

$$\mu_i = \alpha_{\text{sp}[i],\text{pop}[i]} + \beta_{\text{sp}[i],\text{pop}[i]} * \text{year}_i$$

$$\alpha_{\text{sp}[i],\text{pop}[i]} \sim \text{normal}(\mu_{\text{pop}[i]}^\alpha, \sigma_{\text{pop}[i]}^\alpha) \\
\mu_{\text{pop}[i]}^\alpha \sim \text{normal}(\mu_{\text{sp}[i]}^\alpha, \sigma_{\text{sp}[i]}^\alpha)$$

$$\beta{\text{sp}[i],\text{pop}[i]} \sim \text{normal}(\mu_{\text{pop}[i]}^\beta, \sigma_{\text{pop}[i]}^\beta) \\
\mu_{\text{pop}[i]}^\beta \sim \text{normal}(\mu_{\text{sp}[i]}^\beta, \sigma_{\text{sp}[i]}^\beta)$$

*I'm not 100% sure with the notation here.*

@@

So here we refine our question to:

What is trend in mammal species population numbers in Africa, both overall and how does it vary across species and populations?

### Simulating data

Now we start to simulate!

@@

nspecies <- 9 # no. of species
npops_persp <- round(runif(nspecies, 2, 15)) # no. of different populations per species
npops <- sum(npops_persp) # total no. of different populations
nobs_perpop <- round(runif(npops, 10, 40)) # no. of different observations per population

@@

# nested intercepts
mu_species_alpha <- rnorm(n = nspecies, mean = log(200), sd = log(5)) # simulate some species intercepts
sig_species_alpha <- abs(rnorm(n = nspecies, mean = 0, sd = 2))
mu_pop_alpha <- rnorm(n = npops, mean = mu_species_alpha[spid_perpop], sd = sig_species_alpha[spid_perpop]) # population intercepts, nested in species
sig_pop_alpha <- abs(rnorm(n = npops))
alpha <- rnorm(n = npops, mean = mu_pop_alpha, sd = sig_pop_alpha)

# nested slopes
mu_species_beta <- rnorm(n = nspecies, mean = 0, sd = 0.01) # simulate some species slopes
sig_species_beta <- abs(rnorm(n = nspecies, mean = 0, sd = 0.05))
mu_pop_beta <- rnorm(n = npops, mean = mu_species_beta[spid_perpop], sd = sig_species_beta[spid_perpop]) # population slopes, nested in species
sig_pop_beta <- abs(rnorm(n = npops, mean = 0, sd = 0.01))
beta <- rnorm(n = npops, mean = mu_pop_beta, sd = sig_pop_beta)

@@

sigma_obs <- abs(rnorm(1, mean = 0.25, sd = 0.25)) # common observational error

@@

```{r inference sim data}
#| eval: false
#| echo: false

species_intercept <- c(coefs["(Intercept)", "Estimate"], sapply(2:nspecies, function(sp){coefs["(Intercept)", "Estimate"] + coefs[paste0("species", sp), "Estimate"]}))
species_intercept_sd <- c(coefs["(Intercept)", "Std. Error"], sapply(2:nspecies, function(sp){coefs[paste0("species", sp), "Std. Error"]}))

@@

lpi_subset <- lpi_subset[, c("ID", "Genus", "Species", "Country", "Units", paste0("X", 1950:2020))]
lpi_subset$genus_species <- paste(lpi_subset$Genus, lpi_subset$Species)

@@

ggplot(data = lpi_subset_rsh) +
  geom_line(aes(x = year, y = value, 
                group = paste0(genus_species, ID),
                color = genus_species)) +
  theme_bw() +
  labs(x = 'Year', y = 'Observed counts')

ggplot(data = lpi_subset_rsh[lpi_subset_rsh$value != 0,]) +
  geom_line(aes(x = year, y = log(value), 
                group = paste0(genus_species, ID),
                color = genus_species)) +
  theme_bw() +
  labs(x = 'Year', y = 'Observed counts (log-scale)')

@@

runmodel <- FALSE

lpi_subset_rsh <- lpi_subset_rsh[lpi_subset_rsh$value != 0,] # poor way to avoid log(0)
lpi_subset_rsh$year <- lpi_subset_rsh$year-1980
lpi_subset_rsh$logvalue <- log(lpi_subset_rsh$value)
lpi_subset_rsh$ID <- as.character(lpi_subset_rsh$ID)

if(runmodel){
  nestlm <- lm(logvalue ~ genus_species/ID + year * genus_species/ID, data = lpi_subset_rsh)
  saveRDS(nestlm, file = file.path(wd, "output/fit/nestmodel1.rds"))
}else{
  nestlm <- readRDS(file.path(wd, "output/fit/nestmodel1.rds"))
}

@@

```{r}
#| fig-height: 10
#| fig-width: 12
#| eval: true

retrodata <- data.frame(obs = lpi_subset_rsh$logvalue, pred = predict(nestlm),
                        year = lpi_subset_rsh$year, units = lpi_subset_rsh$Units,
                        species = lpi_subset_rsh$genus_species, id = lpi_subset_rsh$ID,
                        country = lpi_subset_rsh$Country)

ggplot(data = retrodata) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey") +
  facet_wrap(~ species, scales = "free") +
  geom_point(aes(x = obs, y = pred, group = id, color = as.character(id)),
             size = 0.5) + 
  theme_bw() + theme(legend.position = 'none') +
  labs(x = 'Observed counts (log-scale)', y = 'Estimated counts (log-scale)')

@@

```{r}
#| fig-height: 8
#| fig-width: 10
#| eval: true

ggplot(data = retrodata[retrodata$species == "Loxodonta africana", ]) +
  facet_wrap(~ country, scales = "free") +
  geom_point(aes(x = year + 1980, y = obs, group = id, color = id),
             size = 1) + 
  geom_line(aes(x = year + 1980, y = pred, group = id, color = id)) +
  theme_bw() + theme(legend.position = 'none') +
  labs(y = 'Counts (log-scale)', x = 'Year')

ggplot(data = retrodata[retrodata$species == "Loxodonta africana", ]) +
  facet_wrap(~ country, scales = "free") +
  geom_point(aes(x = year + 1980, y = exp(obs), group = id, color = id),
             size = 1) + 
  geom_line(aes(x = year + 1980, y = exp(pred), group = id, color = id)) +
  theme_bw() + theme(legend.position = 'none') +
  labs(y = 'Counts', x = 'Year')
